---
title: "Homework 3"
author: "Mia Isaacs"
date: "2024-10-11"
output: github_document
---

# load libraries

```{r}
library(tidyverse)
```

```{r}
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# question 1

```{r}
library(p8105.datasets)
data("ny_noaa")

names(ny_noaa)
```

Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units.

```{r}
noaa_df =  
  janitor::clean_names(ny_noaa) |> 
  mutate(
    year = str_sub(date, 1, 4),
    month = str_sub(date, 6, 7),
    day = str_sub(date, 9, 10),
    prcp = as.numeric(prcp) / 100,
    snow = as.numeric(snow) / 10,
    snwd = as.numeric(snwd) / 10,
    tmax = as.numeric(tmax) / 10,
    tmin = as.numeric(tmin) / 10
  ) |> 
  filter(snow >= 0) |> 
  select(-date)

view(noaa_df)
```

The units for `r prcp` `r snow` and `r snwd` were converted to cm and the unit for `r tmax` and `r tmin` were converted to degrees C.

For snowfall, what are the most commonly observed values? Why?

```{r}

```

The most commonly observed value for `r snow` is (insert inline R code) because most days it does not snow in New York.


Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?

```{r}
noaa_df |> 
  
```


Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.



